An enhancer code compendium of early human neuronal and neural crest development

### 2025-04-08 manuscript.tex
Introduction rewrite.

\title{Deciphering enhancer rules underlying early human neuronal and neural crest development.}
\date{}

@@

Genomes of complex multi-cellular organisms encode all cell types/states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} of that organism in specific non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the arrangement, copy
number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}. Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}. For example, immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}. Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
can be used to study how a genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
interpret both cell intrinsic and extrinsic signals. For example, epigentic
modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
example, early human development.
One such system is early human development, the study of which is especially important for our understanding of human health and disease.
\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their \textit{in vivo} counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}. However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}. In this study, we address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par

During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, achieve different identities along the dorso-ventral (DV) axis of the embryo. These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}. Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}. The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied. Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but detailed descriptions of the enhancer-code of individual cell types are still limited.\par

>>>>

\title{An enhancer code compendium of early human neuronal and neural crest development}
\date{}

@@

Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
Central to the study of the genomic code of enhancers is the paradigm that 
enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020} 
and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par

DNA sequence based deep learning methods\cite{
  alipanahiPredictingSequenceSpecificities2015,
  zhouPredictingEffectsNoncoding2015,
  kelleyBassetLearningRegulatory2016,
  quangDanQHybridConvolutional2016,
  movvaDecipheringRegulatoryDNA2019,
  kelleySequentialRegulatoryActivity2018,
  yuanScBassetSequencebasedModeling2022,
  chenSequencebasedGlobalMap2022,
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  avsecBaseresolutionModelsTranscriptionfactor2021,
  avsecEffectiveGeneExpression2021,
  dealmeidaDeepSTARRPredictsEnhancer2022,
  linderPredictingRNAseqCoverage2025a,
  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
and largely support the enhancer-code paradigm. 
However, these studies are predominantly limited to a small number of cell type or cell states\cite{
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  atakInterpretationAllelespecificChromatin2021,
  avsecEffectiveGeneExpression2021,
  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
  janssensDecodingGeneRegulation2022,
  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par

Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the human genome. 
They can be used to model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. 
Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
can be used to study how the human genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}.\par

@@

Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols 
and how they compare to their \textit{in vivo} counterparts at a transcriptomic level\cite{
  heIntegratedTranscriptomicCell2024,
  xuIntegratedTranscriptomicCell2023}. 
However, such comparisons at the level of the enhancer-code are still limited\cite{
  zenkSinglecellEpigenomicReconstruction2024,
  fleckInferringPerturbingCell2023}.

In this study, we address these limitations by determining the enhancer-codes of cell types and cell states generated by human neural tube organoids, 
a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}
and compare these codes to those of of the head of a four weeks post conception human embryo.
Through these comparisons we confirm the enhancer-code paradigm, 
but also provide a counter examples specifically related to the enhancer-code underlying anterior-posterior patterning.
\par
